---
title: "Final Project"
auther: "Murtaza Yaqubi"
editor: visual
---

### Dataset Introduction:  

This heart disease dataset is acquired from one o f the multispecialty hospitals in India. This dataset contains anonymized patient records related to cardiovascular disease (CVD) risk factors and diagnoses. This dataset consists of 1000 subjects so each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type). This dataset will be useful for generating helpful insights into cardiovasclar diseases and can be used for research purposes. This dataset was obtained from mendeley data where it was first published in April of 2021.

### Lets begin by installing and loading libraries:

```{r}
library(here)
library(readr)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(naniar)
```

### Now lets import the dataset and peek inside it:

```{r}
cardio_df <- read_csv("data/Cardiovascular_Disease_Dataset.csv")
head(cardio_df)
colnames(cardio_df)
dim(cardio_df)
summary(cardio_df)
glimpse(cardio_df)
gg_miss_var(cardio_df)
```

### Research question:

Do younger individuals (e.g., age \< 40) with normal cholesterol levels (within "healthy" range) still face CVD risk if they have elevated blood pressure or ST depression (oldpeak)?

For this analysis I will be using target which is the outcome of interest and is described as abcense or pressence of heart disease and some of the predictors of interest are age, serum cholesterol, resting BP, and old peak. The goal for this analysis is to be able observe if young patients with normal cholesterol but high BP/ST depression have higher CVD rates than expected.

### Breakdown of the analysis: 

The analysis will first isolate younger individuals with clinically normal cholesterol levels to examine their risk of cardiovascular disease (CVD) when other factors like high blood pressure or ST depression are present. Patients will be categorized based on these secondary risk factors, and descriptive analysis will compare CVD prevalence among groups, such as those with normal cholesterol but elevated blood pressure versus those with entirely normal profiles. This step aims to reveal potential associations between non-traditional risk factors and CVD in younger populations.

To validate these findings, statistical tests like chi-square analysis will assess significance, while logistic regression models will quantify the independent and combined effects of blood pressure and ST depression on CVD risk, adjusting for confounders like gender and fasting blood sugar. Visualization techniques, including bar plots and boxplots, will further illustrate risk patterns, making the findings more interpretable. By integrating these methods, the study aims to challenge traditional risk-assessment frameworks and support more nuanced screening approaches for younger individuals at risk of CVD.
